%% The following is a directive for TeXShop to indicate the main file
%%!TEX root = diss.tex

\chapter{Abstract}
%% 345 words

\ac{SMT} solvers are automated decision procedures for domain-specific solving
problems including but not limited to boolean, integer, and real theories,
uninterpreted function theories, array theories, and bit-vector theories.
They have become a main method widely applied in the industry for solving
software and hardware formal verification problems.
Though theorem provers are largely interactive, they are often resorted to for
its powerful induction capability, abstraction capability, and strong proof
management support.
Any moderate-complexity problem will require both methods but switching between
tools requires modelling the same system in each tool.
This introduces huge verification overhead and could be erroneous.

Existing work on the integration of \acs{SMT} solvers and theorem provers
focuses on proof reconstruction for soundness, but lacks emphasis on the
usability and proof efficiency.
To use such integration, the theorem prover term needs to be in a form that is
ready for direct translation.
On the other hand, since proof reconstruction is hard, it is reported that for
some problems the lemma could be time consuming and even unsolvable.
In this thesis, We stress the problem of the usability of such integration.
We propose using reflection for integrating \acs{SMT} solvers into theorem
provers. 
We note that using proof reconstruction to remove the dependency on the
soundness of the \acs{SMT} solver is an orthogonal research direction.

We demonstrate our method using the ACL2 theorem prover and the Z3
\acs{SMT} solver.
To bridge the gap between the untyped first-order logic of ACL2 and the
many-sorted logic of \acs{SMT} solvers, we apply reflection over the ACL2
terms to achieve adding user hypotheses, function expansion, type inference,
term replacement and many other transformations.
By using reflection, the user is freed from manually transforming theorem prover
terms into a shape that is directly translatable.
In addition, soundness of these transformations is easily achieved and does not
require extra proof time, therefore promotes proof efficiency.
The framework is also built to be extensible, allowing new transformations to be
integrated.
We analyze our tool by conducting formal verification of asynchronous circuits
and machine learning convex optimization problems.

% Consider placing version information if you circulate multiple drafts
%\vfill
%\begin{center}
%\begin{sf}
%\fbox{Revision: \today}
%\end{sf}
%\end{center}
